Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-03-31 experiments.tex
Minor fixes

We first compare the ability of the different scoring criteria to
discover the structure that generated the data. For this purpose we
create 100 random 5-node Bayesian network structures with 4
edges and another 100 structures with 7 edges.  The variables were
randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
each network, we generated parameters by two different schemes. The
first scheme exactly matched the assumptions of the BDeu score with
$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
was to generate the parameters independently from a Dirichlet
distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$,
that is, the Jeffreys' prior for the multinomial model. This distribution
was selected instead of the uniform distribution in order to make the
generating structure more identifiable.  For each network (structure +
parameters), we generated 100 data sets of 1000 data vectors, and
studied how different scoring criteria ranked the structure of the
generating network among all the 5-node networks as a function of
(sub)sample size.

@@

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
for all of them. In each experiment, random 10\% of the original
dataset was selected as training data and the rest 90\% as test data.
The training was conducted using dynamic programming based exact structure
learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

@@

\begin{table}
\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
\label{tbl:preds}
\begin{center}
 \begin{tabular}{crrrrr}
    Data &     N &              BDeu &               BIC &               fNML &               qNML \\
\midrule
    Iris &    15 &     \textit{3.61} &              3.56 &   \underline{3.50} &      \textbf{3.47} \\
 PostOpe &    18 &    \textit{10.52} &     \textbf{7.45} &   \underline{8.35} &               8.39 \\
   Ecoli &    34 &     \textit{6.27} &              6.16 &      \textbf{5.56} &   \underline{5.63} \\
   Liver &    35 &     \textit{4.22} &     \textbf{4.01} &               4.10 &   \underline{4.03} \\
    Wine &    36 &    \textit{16.41} &    \textbf{11.27} &              12.07 &  \underline{11.86} \\
   Glass &    44 &     \textit{7.48} &              6.65 &      \textbf{6.37} &   \underline{6.43} \\
 Thyroid &    44 &              2.92 &     \textit{2.93} &      \textbf{2.81} &   \underline{2.83} \\
 HeartSt &    54 &    \textit{14.37} &    \textbf{10.96} &              11.98 &  \underline{11.65} \\
 BreastC &    58 &    \textit{11.21} &     \textbf{9.94} &  \underline{10.34} &              10.37 \\
 HeartHu &    60 &     \textit{9.21} &     \textbf{8.38} &               8.80 &   \underline{8.61} \\
 HeartCl &    62 &    \textit{14.88} &    \textbf{11.76} &              12.48 &  \underline{12.18} \\
 BcWisco &    70 &     \textit{6.41} &     \textbf{5.09} &               5.29 &   \underline{5.28} \\
 Diabete &    77 &     \textit{5.11} &  \underline{4.99} &               5.09 &      \textbf{4.98} \\
 TicTacT &    96 &             10.99 &    \textbf{10.35} &  \underline{10.72} &     \textit{11.04} \\
 Balance &   126 &              7.45 &  \underline{7.42} &      \textbf{7.27} &      \textit{7.73} \\
   Yeast &   149 &              5.67 &     \textit{5.80} &      \textbf{5.48} &   \underline{5.49} \\
 Abalone &   418 &  \underline{3.93} &     \textit{4.00} &      \textbf{3.88} &               3.96 \\
 PageBlo &   548 &  \underline{2.36} &     \textit{2.40} &      \textbf{2.32} &               2.38 \\
 Shuttle &  5800 &  \underline{1.69} &     \textit{1.72} &      \textbf{1.69} &               1.71 \\
   Adult &  6513 &             10.16 &    \textit{10.25} &     \textbf{10.05} &  \underline{10.08} \\
\end{tabular}
\end{center}
\end{table}

@@

Looking at the number of parents for our 20 datasets and sample
sizes (Table~\ref{tbl:nofparams}) again features BIC's preference for
simple models. qNML usually (18/20) beats fNML that performs worst in
14 out of 20 data sets.

\begin{table}
  \caption{Average number of parents per node 
    for different model selection criteria in 20 different datasets.}
\label{tbl:nofparams}
\begin{center}
\begin{tabular}{crrrrr}
    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
\midrule
    Iris &    15 &              0.84 &     \textbf{0.63} &  \textit{0.87} &  \underline{0.82} \\
 PostOpe &    18 &              1.35 &     \textbf{0.11} &  \textit{1.66} &  \underline{1.29} \\
   Ecoli &    34 &              0.83 &     \textbf{0.30} &  \textit{0.96} &  \underline{0.77} \\
   Liver &    35 &              0.60 &     \textbf{0.10} &  \textit{0.88} &  \underline{0.43} \\
    Wine &    36 &     \textit{2.11} &     \textbf{0.75} &           1.86 &  \underline{1.44} \\
   Glass &    44 &              1.32 &     \textbf{0.38} &  \textit{1.41} &  \underline{0.85} \\
 Thyroid &    44 &              0.84 &     \textbf{0.48} &  \textit{0.97} &  \underline{0.66} \\
 HeartSt &    54 &              1.81 &     \textbf{0.56} &  \textit{2.06} &  \underline{1.50} \\
 BreastC &    58 &  \underline{1.27} &     \textbf{0.42} &  \textit{1.71} &              1.39 \\
 HeartHu &    60 &  \underline{0.93} &     \textbf{0.56} &  \textit{1.55} &              1.04 \\
 HeartCl &    62 &              1.69 &     \textbf{0.41} &  \textit{1.96} &  \underline{1.33} \\
 BcWisco &    70 &     \textit{2.02} &     \textbf{0.76} &           1.96 &  \underline{1.14} \\
 Diabete &    77 &  \underline{0.55} &     \textbf{0.23} &  \textit{1.04} &              0.58 \\
 TicTacT &    96 &  \underline{1.60} &     \textbf{0.21} &           2.32 &     \textit{2.47} \\
 Balance &   126 &     \textbf{0.06} &  \underline{0.15} &           0.71 &     \textit{1.18} \\
   Yeast &   149 &              0.54 &     \textbf{0.14} &  \textit{0.78} &  \underline{0.45} \\
 Abalone &   418 &              1.20 &     \textbf{0.78} &  \textit{1.49} &  \underline{0.93} \\
 PageBlo &   548 &     \textit{1.67} &     \textbf{0.38} &           1.17 &  \underline{0.49} \\
 Shuttle &  5800 &     \textit{2.02} &     \textbf{1.05} &           1.84 &  \underline{1.14} \\
   Adult &  6513 &  \underline{1.34} &     \textbf{1.11} &  \textit{1.59} &              1.36 \\
\end{tabular}
\end{center}
\end{table}

>>>>

We first compare the ability of the different scoring criteria to
discover the structure that generated the data. For this purpose we
create 100 random 5-node Bayesian network structures with 4
edges and another 100 structures with 7 edges.  The variables were
randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
each network, we generated parameters by two different schemes. The
first scheme exactly matched the assumptions of the BDeu score with
$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
was to generate the parameters independently from a Dirichlet
distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
that is the Jeffreys' prior for the multinomial model. This distribution
was selected instead of the uniform distribution in order to make the
generating structure more identifiable.  For each network (structure +
parameters), we generated 100 data sets of 1000 data vectors, and
studied how different scoring criteria ranked the structure of the
generating network among all the 5-node networks as a function of
(sub)sample size.

@@

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
for all of them. In each experiment, random 10\% of the original
data set was selected as training data and the rest 90\% as test data.
The training was conducted using dynamic programming based exact structure
learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

@@

\begin{table}
\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different data sets.}
\label{tbl:preds}
\begin{center}
 \begin{tabular}{crrrrr}
    Data &     N &              BDeu &               BIC &               fNML &               qNML \\
\midrule
    Iris &    15 &     \textit{3.61} &              3.56 &   \underline{3.50} &      \textbf{3.47} \\
 PostOpe &    18 &    \textit{10.52} &     \textbf{7.45} &   \underline{8.35} &               8.39 \\
   Ecoli &    34 &     \textit{6.27} &              6.16 &      \textbf{5.56} &   \underline{5.63} \\
   Liver &    35 &     \textit{4.22} &     \textbf{4.01} &               4.10 &   \underline{4.03} \\
    Wine &    36 &    \textit{16.41} &    \textbf{11.27} &              12.07 &  \underline{11.86} \\
   Glass &    44 &     \textit{7.48} &              6.65 &      \textbf{6.37} &   \underline{6.43} \\
 Thyroid &    44 &              2.92 &     \textit{2.93} &      \textbf{2.81} &   \underline{2.83} \\
 HeartSt &    54 &    \textit{14.37} &    \textbf{10.96} &              11.98 &  \underline{11.65} \\
 BreastC &    58 &    \textit{11.21} &     \textbf{9.94} &  \underline{10.34} &              10.37 \\
 HeartHu &    60 &     \textit{9.21} &     \textbf{8.38} &               8.80 &   \underline{8.61} \\
 HeartCl &    62 &    \textit{14.88} &    \textbf{11.76} &              12.48 &  \underline{12.18} \\
 BcWisco &    70 &     \textit{6.41} &     \textbf{5.09} &               5.29 &   \underline{5.28} \\
 Diabete &    77 &     \textit{5.11} &  \underline{4.99} &               5.09 &      \textbf{4.98} \\
 TicTacT &    96 &             10.99 &    \textbf{10.35} &  \underline{10.72} &     \textit{11.04} \\
 Balance &   126 &              7.45 &  \underline{7.42} &      \textbf{7.27} &      \textit{7.73} \\
   Yeast &   149 &              5.67 &     \textit{5.80} &      \textbf{5.48} &   \underline{5.49} \\
 Abalone &   418 &  \underline{3.93} &     \textit{4.00} &      \textbf{3.88} &               3.96 \\
 PageBlo &   548 &  \underline{2.36} &     \textit{2.40} &      \textbf{2.32} &               2.38 \\
 Shuttle &  5800 &  \underline{1.69} &     \textit{1.72} &      \textbf{1.69} &               1.71 \\
   Adult &  6513 &             10.16 &    \textit{10.25} &     \textbf{10.05} &  \underline{10.08} \\
\end{tabular}
\end{center}
\end{table}

@@

Looking at the number of parents for our 20 data sets and sample
sizes (Table~\ref{tbl:nofparams}) again features BIC's preference for
simple models. qNML usually (18/20) beats fNML that performs worst in
14 out of 20 data sets.

\begin{table}
  \caption{Average number of parents per node 
    for different model selection criteria in 20 different data sets.}
\label{tbl:nofparams}
\begin{center}
\begin{tabular}{crrrrr}
    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
\midrule
    Iris &    15 &              0.84 &     \textbf{0.63} &  \textit{0.87} &  \underline{0.82} \\
 PostOpe &    18 &              1.35 &     \textbf{0.11} &  \textit{1.66} &  \underline{1.29} \\
   Ecoli &    34 &              0.83 &     \textbf{0.30} &  \textit{0.96} &  \underline{0.77} \\
   Liver &    35 &              0.60 &     \textbf{0.10} &  \textit{0.88} &  \underline{0.43} \\
    Wine &    36 &     \textit{2.11} &     \textbf{0.75} &           1.86 &  \underline{1.44} \\
   Glass &    44 &              1.32 &     \textbf{0.38} &  \textit{1.41} &  \underline{0.85} \\
 Thyroid &    44 &              0.84 &     \textbf{0.48} &  \textit{0.97} &  \underline{0.66} \\
 HeartSt &    54 &              1.81 &     \textbf{0.56} &  \textit{2.06} &  \underline{1.50} \\
 BreastC &    58 &  \underline{1.27} &     \textbf{0.42} &  \textit{1.71} &              1.39 \\
 HeartHu &    60 &  \underline{0.93} &     \textbf{0.56} &  \textit{1.55} &              1.04 \\
 HeartCl &    62 &              1.69 &     \textbf{0.41} &  \textit{1.96} &  \underline{1.33} \\
 BcWisco &    70 &     \textit{2.02} &     \textbf{0.76} &           1.96 &  \underline{1.14} \\
 Diabete &    77 &  \underline{0.55} &     \textbf{0.23} &  \textit{1.04} &              0.58 \\
 TicTacT &    96 &  \underline{1.60} &     \textbf{0.21} &           2.32 &     \textit{2.47} \\
 Balance &   126 &     \textbf{0.06} &  \underline{0.15} &           0.71 &     \textit{1.18} \\
   Yeast &   149 &              0.54 &     \textbf{0.14} &  \textit{0.78} &  \underline{0.45} \\
 Abalone &   418 &              1.20 &     \textbf{0.78} &  \textit{1.49} &  \underline{0.93} \\
 PageBlo &   548 &     \textit{1.67} &     \textbf{0.38} &           1.17 &  \underline{0.49} \\
 Shuttle &  5800 &     \textit{2.02} &     \textbf{1.05} &           1.84 &  \underline{1.14} \\
   Adult &  6513 &  \underline{1.34} &     \textbf{1.11} &  \textit{1.59} &              1.36 \\
\end{tabular}
\end{center}
\end{table}

### 2017-03-31 intro.tex
Minor fixes

The information-theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion, but its exact calculation is likely to be
prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike the fNML,
the qNML gives equal scores to structures that encode the same
independence statements. Like other common model selection criteria,
qNML is also consistent.

We next briefly introduce Bayesian networks and review the BDeu and
fNML criteria and then introduce the qNML criterion.  We also
summarize the results for 20 datasets to back up our claim that qNML
yields parsimonious models with good predictive capabilities. The
experiments with artificial data demonstrate the capability to give
high scores for the data generating model with relatively small sample
sizes.

>>>>

The information-theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be a potential candidate
for a good criterion, but its exact calculation is likely to be
prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike the fNML,
the qNML gives equal scores to structures that encode the same
independence statements. Like other common model selection criteria,
qNML is also consistent.

We next briefly introduce Bayesian networks and review the BDeu and
fNML criteria and then introduce the qNML criterion.  We also
summarize the results for 20 data sets to back up our claim that qNML
yields parsimonious models with good predictive capabilities. The
experiments with artificial data demonstrate the capability to give
high scores for the data generating model with relatively small sample
sizes.

### 2017-03-31 qnml.tex
Minor fixes

qNML yields equal scores for network structures that encode same sets
of independencies. Verma and Pearl~\cite{Verm90} showed that the
equivalent networks are exactly those which a) are same when directed
arcs are substituted by undirected ones and b) which have same
\textit{V-structures}, i.e. the variable triplets $(A,B,C)$ where both
$A$ and $B$ are parents of $C$, but there is no arc between $A$ and
$B$ (in either direction).  Later, Chickering~\cite{Chick95} showed
that all the equivalent network structures, and only those structures,
can be reached from each other by reversing, one by one, the so-called
\textit{covered arcs}, i.e. the arcs from node $A$ to $B$, for which
$B$'s parents other than $A$ are exactly  $A$'s parents
($G_B=\{A\}\cup G_A$).

>>>>

qNML yields equal scores for network structures that encode same sets
of independencies. Verma and Pearl~\cite{Verm90} showed that the
equivalent networks are exactly those which a) are the same when directed
arcs are substituted by undirected ones and b) which have the same
\textit{V-structures}, i.e. the variable triplets $(A,B,C)$ where both
$A$ and $B$ are parents of $C$, but there is no arc between $A$ and
$B$ (in either direction).  Later, Chickering~\cite{Chick95} showed
that all the equivalent network structures, and only those structures,
can be reached from each other by reversing, one by one, the so-called
\textit{covered arcs}, i.e. the arcs from node $A$ to $B$, for which
$B$'s parents other than $A$ are exactly  $A$'s parents
($G_B=\{A\}\cup G_A$).
